Towards Practical First-Order Model Counting

### 2025-01-18 doc/paper/paper.tex
Use todonotes

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{booktabs}
\usepackage{pifont}
\usepackage{siunitx}
\usepackage{tikz}
\usepackage{microtype}

@@

% TODO (for the entire paper):
% * finish add DOIs or URLs to (non-book) references (and reformatting indentation) (and check reference formatting)
% * sentence vs formula: be consistent and not confusing
% * 15 pages excluding references

@@

\section{Introduction}
% TODO: We would like to clarify that the main contribution of this work
% consists of everything needed to complement recursive function definitions
% with the necessary base cases. This process includes identifying the base
% cases and their corresponding formulas, transforming them (including applying
% a new smoothing procedure), and recursively reusing Crane. C++ code
% generation, although relatively straightforward, is crucial for the usability
% of the algorithm

% TODO: focus more on current work

@@

% TODO: don't assume that the reader knows FOMC and FOMC algorithms

% TODO: It would be nice to give the necessary introduction to FOKC, or CRANE
% (an intuitive explanation early in the paper)

% TODO: The principal motivation of a knowledge compilation-based approach to
% FOMC comes from its ability to find polynomial-time solutions, where the
% polynomial has a lower degree compared to, e.g., FastWFOMC (Dilkas and Belle,
% 2023). The ability of the algorithm to mathematically describe recursive
% solutions is also noteworthy as the only other available alternative is to
% guess them (Barvı́nek, van Bremen, Wang, Zelezný, Kuzelka, ILP 2021).
FOMC algorithms are diverse, with approaches ranging from \emph{first-order
  knowledge compilation} (FOKC) to local
search~\cite{DBLP:journals/pvldb/NiuRDS11}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime
approximation~\cite{DBLP:conf/ijcai/BremenK20}. Among these, FOKC-based
algorithms are particularly prominent, transforming formulas into structured
representations such as circuits or graphs. Notable examples include
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its successor
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. Another important algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, uses cell enumeration as its
foundation.

@@

\begin{example}
  Skolemization transforms
  \begin{equation}\label[formula]{eq:skolemizationinitial}
    \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. } P(x, y)
  \end{equation}
  into
  \begin{equation}\label[formula]{eq:skolemization}
    \begin{aligned}
      (\forall x \in \Gamma\text{. } &Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &Z(x) \lor \neg P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. } &S(x) \lor Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &S(x) \lor \neg P(x, y)).
    \end{aligned}
  \end{equation}
  We will see how, with suitable weights on the new predicates $S/1$ and $Z/1$,
  the WFOMC of \cref{eq:skolemization} is equal to the FOMC of
  \cref{eq:skolemizationinitial}.
\end{example}
% TODO: make sure to revisit this in the future

@@

% TODO: shorten the part about the three logics

@@

While this work focuses on FOMC, we still define the weighted variant of the
problem as Skolemization relies on weights even for unweighted FOMC.
% TODO: Even though we are evaluating the algorithms on FOMC benchmarks, for
% sentences with existential quantifiers, computing FOMC using Gantry requires
% the use of WFOMC. For such sentences, Skolemization introduces predicates with
% non-unary weights that must be accounted for to compute the correct model
% count.

@@

% TODO: The algorithm’s recursive nature and its interaction with CRANE could be
% explained more clearly

% TODO: more emphasis on the motivation for each part of GANTRY (e.g., how
% FindBaseCases addresses key issues in CRANE) would add depth to the
% contribution.

% TODO: this section should be expanded by 1 page ("a clear discussion of the
% innovation aspect and technical parts")

@@

% TODO (somewhere): Recall that each equation is either the definition (did I
% define what a definition is?) of a function or the definition of a base case
% of a function. (I could put this in the preliminaries as definitions and then
% reference them below)

@@

% TODO: move this example right after the definition of WFOMC (or near the
% definition of a model)
\begin{example}[Counting bijections]\label{example:overall}
  Consider the following formula (previously examined by Dilkas and
  Belle~\cite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a bijection
  between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y))\land{}\\
      (\forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z)\land{}\\
      (\forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z).
    \end{gathered}
  \]
  We specifically examine the first solution returned by \Cranetwo{} for this
  formula.

@@

\subsection{Identifying a Sufficient Set of Base Cases}\label{sec:identifying}
% TODO: prove correctness?

@@

% TODO Experiments could be expanded by 1 page ("a more thorough and independent
% practical assessment")

% TODO: this section "would benefit from cleaner reasoning"

% TODO: The difference between Gantry-Greedy and Gantry-BFS is in how the
% algorithm chooses which compilation rule to apply to a formula. The former
% uses greedy search: there is a list of rules, and the first applicable rule is
% the one that gets used, disregarding all the others. The latter uses a
% combination of greedy and breadth-first search (BFS). That is, each
% compilation rule is identified as either greedy or non-greedy. Greedy rules
% are applied as soon as possible at any stage of the compilation process. BFS
% is executed over all applicable non-greedy rules, identifying the solution
% that can be constructed using the smallest number of such rules.

@@

\subsection{Benchmarks}
% TODO: more benchmarks from NOT my work
% TODO: make sure it's clear what the 'raw' instances are
% TODO: don't emphasise that the benchmarks come from my previous work

@@

\begin{figure*}[t]
  \centering
  \includegraphics{plot.pdf}
  \caption{The runtime of the algorithms as a function of the domain size. Note
    that both axes are on a logarithmic scale.}\label{fig:plot}
\end{figure*}
% TODO: shrink/restructure to fit into the margins

@@

% TODO: On the ’friends’ and ’functions’ benchmarks, ForcLift runs until the
% model count exceeds 231 − 1.

% TODO: We are not aware of any formulas on which Gantry scales worse compared
% to either ForcLift or FastWFOMC. The one advantage that FastWFOMC has over
% Gantry is its support for counting quantifiers.

% TODO: Regarding programming languages and accuracy, we verified that the
% answers match for smaller domain sizes. Also, although written in different
% languages, both Gantry and FastWFOMC use the GNU Multiple Precision Arithmetic
% Library.

>>>>

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{booktabs}
\usepackage{pifont}
\usepackage{siunitx}
\usepackage{tikz}
\usepackage{microtype}
\usepackage[backgroundcolor=lightgray]{todonotes}

@@

@@

\todo[inline,caption={}]{
  For the entire paper:
  \begin{itemize}
    \item Finish adding DOIs or URLs to (non-book) references (and reformatting
          indentation) (and check reference formatting)
    \item Sentence vs formula: be consistent and not confusing
    \item 15 pages excluding references
    \item Go through the rest of formatting instructions (in GTD)
  \end{itemize}
}
\section{Introduction}

\todo[inline,caption={}]{
  \begin{itemize}
    \item We would like to clarify that the main contribution of this work
          consists of everything needed to complement recursive function
          definitions with the necessary base cases. This process includes
          identifying the base cases and their corresponding formulas,
          transforming them (including applying a new smoothing procedure), and
          recursively reusing Crane. C++ code generation, although relatively
          straightforward, is crucial for the usability of the algorithm.
    \item Focus more on current work
  \end{itemize}
}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Don't assume that the reader knows FOMC and FOMC algorithms
    \item It would be nice to give the necessary introduction to FOKC, or
          \textsc{Crane} (an intuitive explanation early in the paper)
    \item The principal motivation of a knowledge compilation-based approach to
          FOMC comes from its ability to find polynomial-time solutions, where
          the polynomial has a lower degree compared to, e.g.,
          \textsc{FastWFOMC}~\cite{DBLP:conf/kr/DilkasB23}. The ability of the
          algorithm to mathematically describe recursive solutions is also
          noteworthy as the only other available alternative is to guess
          them~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.
  \end{itemize}
}

FOMC algorithms are diverse, with approaches ranging from \emph{first-order
  knowledge compilation} (FOKC) to local
search~\cite{DBLP:journals/pvldb/NiuRDS11}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime
approximation~\cite{DBLP:conf/ijcai/BremenK20}. Among these, FOKC-based
algorithms are particularly prominent, transforming formulas into structured
representations such as circuits or graphs. Notable examples include
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its successor
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. Another important algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, uses cell enumeration as its
foundation.

@@

\begin{example}
  Skolemization transforms
  \begin{equation}\label[formula]{eq:skolemizationinitial}
    \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. } P(x, y)
  \end{equation}
  into
  \begin{equation}\label[formula]{eq:skolemization}
    \begin{aligned}
      (\forall x \in \Gamma\text{. } &Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &Z(x) \lor \neg P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. } &S(x) \lor Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &S(x) \lor \neg P(x, y)).
    \end{aligned}
  \end{equation}
  We will see how, with suitable weights on the new predicates $S/1$ and $Z/1$,
  the WFOMC of \cref{eq:skolemization} is equal to the FOMC of
  \cref{eq:skolemizationinitial}.
\end{example}

\todo[inline]{Make sure to revisit this in the future}

@@

\todo[inline]{Shorten the part about the three logics}

@@

While this work focuses on FOMC, we still define the weighted variant of the
problem as Skolemization relies on weights even for unweighted FOMC.

\todo[inline]{Even though we are evaluating the algorithms on FOMC benchmarks,
  for sentences with existential quantifiers, computing FOMC using \Cranetwo{}
  requires the use of WFOMC. For such sentences, Skolemization introduces
  predicates with non-unary weights that must be accounted for to compute the
  correct model count.}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item The algorithm’s recursive nature and its interaction with
          \textsc{Crane} could be explained more clearly
    \item More emphasis on the motivation for each part of \Cranetwo{} (e.g.,
          how \FindBaseCases addresses key issues in \textsc{Crane}) would add
          depth to the contribution
    \item this section should be expanded by 1 page (``a clear discussion of the
          innovation aspect and technical parts'')
  \end{itemize}
}

@@

\todo[inline]{Somewhere: Recall that each equation is either the definition (did
  I define what a definition is?) of a function or the definition of a base case
  of a function. (I could put this in the preliminaries as definitions and then
  reference them below)}

@@

\todo[inline]{Move this example right after the definition of WFOMC (or near the
  definition of a model)}

\begin{example}[Counting bijections]\label{example:overall}
  Consider the following formula (previously examined by Dilkas and
  Belle~\cite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a bijection
  between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y))\land{}\\
      (\forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z)\land{}\\
      (\forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z).
    \end{gathered}
  \]
  We specifically examine the first solution returned by \Cranetwo{} for this
  formula.

@@

\subsection{Identifying a Sufficient Set of Base Cases}\label{sec:identifying}

\todo[inline]{Prove correctness?}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Experiments could be expanded by 1 page (``a more thorough and
          independent practical assessment'')
    \item The difference between \Cranegreedy{} and \Cranebfs{} is in how the
          algorithm chooses which compilation rule to apply to a formula. The
          former uses greedy search: there is a list of rules, and the first
          applicable rule is the one that gets used, disregarding all the
          others. The latter uses a combination of greedy and breadth-first
          search (BFS). That is, each compilation rule is identified as either
          greedy or non-greedy. Greedy rules are applied as soon as possible at
          any stage of the compilation process. BFS is executed over all
          applicable non-greedy rules, identifying the solution that can be
          constructed using the smallest number of such rules.
  \end{itemize}
}

@@

\subsection{Benchmarks}

\todo[inline,caption={}]{
  \begin{itemize}
    \item More benchmarks from NOT my work
    \item Make sure it's clear what the `raw' instances are
    \item Don't emphasise that the benchmarks come from my previous work
  \end{itemize}
}

@@

\begin{figure*}[t]
  \centering
  \includegraphics{plot.pdf}
  \caption{The runtime of the algorithms as a function of the domain size. Note
    that both axes are on a logarithmic scale.}\label{fig:plot}
\end{figure*}

\todo[inline]{Shrink/restructure to fit into the margins}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item On the `friends' and `functions' benchmarks, \textsc{ForcLift} runs
          until the model count exceeds $2^{31}-1$.
    \item We are not aware of any formulas on which \Cranetwo{} scales worse
          compared to either \textsc{ForcLift} or \textsc{FastWFOMC}. The one
          advantage that \textsc{FastWFOMC} has over \Cranetwo{} is its support
          for counting quantifiers.
    \item Regarding programming languages and accuracy, we verified that the
          answers match for smaller domain sizes. Also, although written in
          different languages, both \Cranetwo{} and \textsc{FastWFOMC} use the
          GNU Multiple Precision Arithmetic Library.
  \end{itemize}
}
